|

### 2025-09-05 manuscript/paper.Rmd
updated.

---
title: |
  | \large Assessing Ranking Fidelity of Competition Structures via Weighted Mutual Information
author: | 
  | \large Zach Culp \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Josie Peterburs \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Ryan McShane \vspace{-1.1mm}
  | \normalsize University of Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  |	\normalsize  \vspace{-1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`email`](mailto:ypu@something.edu) \vspace{-1mm}
  | \normalsize [`zculp@luc.edu`](mailto:zculp@luc.edu) \vspace{-1mm}
  | \normalsize [`jpeterburs@luc.edu`](mailto:jpeterburs@luc.edu) \vspace{-1mm}
  | \normalsize [`rmcshane@uchicago.edu`](mailto:rmcshane@uchicago.edu) \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: | 
 | We evaluate multiple competitions structures to assess their effectiveness in accurately reproducing the true underlying ranking of participating teams. Using repeated simulations for each structure, we compare teams’ true performance ranks to their simulated outcomes. To quantify the accuracy of each format, we calculate weighted mutual information between true and simulated rankings across a large number of replicates. "Salient" weights are applied to look at only the top n teams/competitors in a given competition structure. This information-theoretic approach allows us to objectively compare formats and identify which structures most effectively preserve rank order. Results are visualized through comparative plots, providing clear insights into the trade-offs and strengths of each competition design.
 | 
  \vspace{2mm}
  | Keywords: tournament structures, mutual information
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
editor_options: 
  markdown: 
    wrap: 72
---

@@

For example, in a 4 team tournament, $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ = (3, 2, 1, 4) indicates that the team with index 3 (i.e. the true third best team) won the
tournament, the true second best team finished second, the true best team finished third and the true 4th team finished 4th in that particular tournament. If $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ = (1, 2, 3, 4), this means that the random outcome of the tournament was the same as the true ordering of the teams.

@@

However, because the number of teams that need to be accurately ordered differs based on the choice of salient ranks, a second weighting function, $m(r)$, is used to indicate which ranks are salient.  We call these the "salient weights" to differentiate from the other weighting function.  As an example, for a four-team tournament, if the organizer decides that only the winner is important, then $m(r) = \{1,0,0,0\}$ (i.e. essentially measuring effectivity).  While we only consider indicators (i.e. 0's or 1's) in the salient weights vector, there is no reason why you couldn't weight the ranks by relative levels of importance.  For instance, letting $m(r) = \{3,1,0,0\}$ indicates that the salient ranks are 1 and 2, but getting the true best competitor as the winner is more important than ranking the true second best competitor second.

@@

$$
I(X;Y)=\sum_{r}\sum_{\hat{r}} w(r,\hat{r}) * m(r) * p(r,\hat{r})\log \frac {p(r,\hat{r})}{p(r)p(\hat{r})}
$$

Because this mutual information is unitless, the formula can be
standardized to be between 0 and 1 using the complement of Rajski's
distance (@Rajski1961): $$
1 - (1 - \frac{I(X,Y)}{\text{max\{}H(X), H(Y)\}})
$$

where $H(X,Y)$ is entropy given by:

$$
max\{H(X), H(Y)\} = \sum_{r}  p(r) \log p(r) = n! \log n!
$$

The complement is taken so that the more accurate tournaments are closer
to 1 and the less accurate are closer to 0.

@@

@@

![Inofrmartion Curves \label{fig:fig1}](images/Normal_Information_Curve_Graph.png){width="408"}

Figure 1 \ref{fig:fig1}

>>>>

---
title: |
  | \large Assessing Ranking Fidelity of Competition Structures via Weighted Mutual Information
author: | 
  | \large Zach Culp \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Josie Peterburs \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Ryan McShane \vspace{-1.1mm}
  | \normalsize University of Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  |	\normalsize  \vspace{-1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`email`](mailto:ypu@something.edu) \vspace{-1mm}
  | \normalsize [`zculp@luc.edu`](mailto:zculp@luc.edu) \vspace{-1mm}
  | \normalsize [`jpeterburs@luc.edu`](mailto:jpeterburs@luc.edu) \vspace{-1mm}
  | \normalsize [`rmcshane@uchicago.edu`](mailto:rmcshane@uchicago.edu) \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: | 
 | We evaluate multiple competitions structures to assess their effectiveness in accurately reproducing the true underlying ranking of participating teams. Using repeated simulations for each structure, we compare teams’ true performance ranks to their simulated outcomes. To quantify the accuracy of each format, we calculate weighted mutual information between true and simulated rankings across a large number of replicates. "Salient" weights are applied to look at only the top i out of n total teams/competitors in a given competition structure. This information-theoretic approach allows us to objectively compare formats and identify which structures most effectively preserve rank order. Results are visualized through comparative plots, providing clear insights into the trade-offs and strengths of each competition design.
 | 
  \vspace{2mm}
  | Keywords: tournament structures, mutual information
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
editor_options: 
  markdown: 
    wrap: 72
---

@@

For example, in a $n$ = 4 team tournament, $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ = (3, 2, 1, 4) indicates that the team with index 3 (i.e. the true third best team) won the
tournament, the true second best team finished second, the true best team finished third and the true 4th team finished 4th in that particular tournament. If $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ = (1, 2, 3, 4), this means that the random outcome of the tournament was the same as the true ordering of the teams.

@@

However, because the number of teams that need to be accurately ordered differs based on the choice of salient ranks, a second set of weights is defined, $m$, is used to indicate which ranks are salient.  We call these the "salient weights" to differentiate from the other weighting function.  As an example, for a four-team tournament, if the organizer decides that only the winner is important, then $m = \{1,0,0,0\}$ (i.e. essentially measuring effectivity).  While here we only consider indicators (i.e. 0's or 1's) in the salient weights vector, there is no reason why you couldn't weight the ranks by relative levels of importance.  For instance, letting $m = \{3,1,0,0\}$ indicates that the salient ranks are 1 and 2, but getting the true best competitor as the winner is more important than ranking the true second best competitor second.

@@

$$
wI_m(r, \hat{r})=\sum_{r}\sum_{\hat{r}} w(r,\hat{r}) * m * p(r,\hat{r})\log \frac {p(r,\hat{r})}{p(r)p(\hat{r})}
$$

Because this mutual information is unitless, the formula can be standardized to be between 0 and 1 as follows: 
$$
\frac{wI_m(r,\hat{r})}{\text{max\{}H(r), H(\hat{r})\}}
$$

where $H(.)$ is the usual Shannon entropy.  Competition structures with higher levels of efficacy are indicated when our measure is closer to 1.  Likewise, competition structures that are near 0 by our proposed metric indicate low levels of efficacy.

<!-- $$ -->
<!-- max\{H(X), H(Y)\} = \sum_{r}  p(r) \log p(r) = n! \log n! -->
<!-- $$ -->

@@

What's interesting: 
1. Single and double are really close at 1 or 2, but they separate past 3.  
2.  The black curve crosses the yellow and green curve. 
3.  The yellow and black curves actually increases towards the end.

We examined the following structure: 
 - Single elimination, usual seeding (1 vs 8, 4 vs 5, 3 vs 6, 2 vs 7)
 - Double elimination

@@

![Zach add a caption \label{fig:fig1}](images/Normal_Information_Curve_Graph.png){width="408"}

Figure \ref{fig:fig1}
